Physcraper: A Python package for continually updated gene trees

### 2020-06-23 physcraper_ms.Rmd
blast command tools

- The next step is to identify the search taxon within the reference taxonomy. The search taxon will be used to constraint the DNA sequence search on the nucleotide database within that taxonomic group. Because we are using the NCBI nucleotide database, by default the reference taxonomy is the NCBI taxonomy. The search taxon can be provided by the user. If none is provided, then the search taxon is identified as the Most Recent Common Ancestor (MRCA) of the
  matched taxa belonging to the ingroup in the tree, that is also a named clade in the reference taxonomy. This is known as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred to as the Least Inclusive Common Ancestral Taxon - LICA).
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the reference taxonomy.
  To automatically identify the MRCAT of a group of taxon names, we make use of the OToL taxonomy tool (<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>).

  In the case that only one taxon from the ingroup is matched in both the tree and alignment, the MRCAT
  for that single taxon would be determined as ***HOW?***
  <!-- ***Physcraper does not take into account the focal clade from Open Tree. Because that is arbitrarily assigned or not by a curator. The focal clade is a search MRCA more than a taxonomic MRCA, physcraper's default is to get the ingroup from Otol, which is not the focal group.-->
  If the search taxon is provided by the user, it can be a more inclusive
  clade, if the user wants to perform a wider search, outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order of the ingroup. It can also be a less inclusive clade, if the user only
  wants to focus on enriching a particular clade/region within the ingroup.
- The Basic local alignment search tool, BLAST [@altschul1990basic; altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the remaining sequences on the alignment.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- A pairwise all-against-all BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding ide numbers (accesion numbers in the case of the GenBank database). This information will be
  used later to store the whole sequences into a new library.
- The DNA sequence similarity search can be done on a local database that is easily
  setup by the user. In this case, the BLASTn algorithm is used to performs the similarity search.
- The search can also be performed remotely, on the NCBI database. In this case, the
  bioPython BLAST algorithm is used to perform the similarity search.
- Matched sequences below an e-value, percentage similarity, and outside a minimum
  and maximum length threshold are discarded. This filtering leaves out genomic sequences.
  All acepted sequences are asigned an internal identifier, and are further filtered.
- Because the original alignments usually lack database id numbers, a filtering
  step is needed. Accepted sequences that belong to the
  same taxon of the query sequence, and that are either identical or shorter than
  the original sequence are discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered further for analysis.
- Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them to, for example, investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
  <!--***How does pyphlawd, or phylota does the exemplar per species choosing?***-->
- Reverse complement sequences are identified and translated.
- Users can choose to perform a more "cycles" of sequence similarity search, by blasting the newly found sequences. This can be done iteratively, but by default only sequences in the alignment are blasted. ***Is there an argument to control the number of cycles of blast searches with new sequences?***
- Accepted sequences are downloaded in full, and stored as a local database in a directory that is globally accesible (physcraper/taxonomy), so they are accesible for further runs.
- A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user.

>>>>

- The next step is to identify the search taxon within the reference taxonomy. The search taxon will be used to constraint the DNA sequence search on the nucleotide database within that taxonomic group. Because we are using the NCBI nucleotide database, by default the reference taxonomy is the NCBI taxonomy. The search taxon can be provided by the user. If none is provided, then the search taxon is identified as the Most Recent Common Ancestor (MRCA) of the
  matched taxa belonging to the ingroup in the tree, that is also a named clade in the reference taxonomy. This is known as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred in the literature as the Least Inclusive Common Ancestral Taxon - LICA).
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the reference taxonomy.
  To automatically identify the MRCAT of a group of taxon names, we make use of the OToL taxonomy tool (<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>).

  In the case that only one taxon from the ingroup is matched in both the tree and alignment, the MRCAT
  for that single taxon would be determined as ***HOW?***
  <!-- ***Physcraper does not take into account the focal clade from Open Tree. Because that is arbitrarily assigned or not by a curator. The focal clade is a search MRCA more than a taxonomic MRCA, physcraper's default is to get the ingroup from Otol, which is not the focal group.-->
  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the remaining sequences on the alignment.
  The BLAST command line tools [@camacho2009blast] are used for both web-based and local-database searches.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- A pairwise alignment-against-all BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accesion numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the physcraper folder, allowing for secondary analyses to run significantly faster.
- The DNA sequence similarity search can be done on a local database that is easily
  setup by the user. In this case, the BLASTn algorithm is used to performs the similarity search.
- The search can also be performed remotely, on the NCBI database. In this case, the
  bioPython BLAST algorithm is used to perform the similarity search.
- Matched sequences below an e-value, percentage similarity, and outside a minimum
  and maximum length threshold are discarded. This filtering leaves out genomic sequences.
  All acepted sequences are asigned an internal identifier, and are further filtered.
- Because the original alignments usually lack database id numbers, a filtering
  step is needed. Accepted sequences that belong to the
  same taxon of the query sequence, and that are either identical or shorter than
  the original sequence are discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered further for analysis.
- Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them to, for example, investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
  <!--***How does pyphlawd, or phylota does the exemplar per species choosing?***-->
- Reverse complement sequences are identified and translated.
- Users can choose to perform a more "cycles" of sequence similarity search, by blasting the newly found sequences. This can be done iteratively, but by default only sequences in the alignment are blasted. ***Is there an argument to control the number of cycles of blast searches with new sequences?***
- Accepted sequences are downloaded in full, and stored as a local database in a directory that is globally accesible (physcraper/taxonomy), so they are accesible for further runs.
- A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user.
